Local Spectral Clustering of Density Upper Level Sets

### 2019-11-21 arxiv/local_spectral_density_clustering_arxiv.tex
arxived version

\begin{abstract}
We analyze the Personalized PageRank (PPR) algorithm, a local spectral method
for clustering, which extracts clusters using locally-biased random walks around
a given seed node.  In contrast to previous work, we adopt a classical
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters, that satisfy a set of natural geometric
conditions. We also show a converse result, that PPR can fail to recover
geometrically poorly-conditioned density clusters, even asymptotically. Finally,
we provide empirical support our theory.
\end{abstract}
\end{center}

@@

First, we define the PPR vector $\pbf_v = \pbf(v,\alpha;G_{n,r})$, based on 
a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to
be the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf_v = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in
position $v$ and 0 elsewhere).

@@

Let \smash{$\Cbb_f(\lambda)$} denote the connected components of the density
upper level set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster
\smash{$\Cset \in \Cbb_f(\lambda)$}, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the
\emph{empirical density cluster}. The size of the symmetric set difference
between estimated and empirical cluster is a commonly used metric to quantify
cluster estimation error \citep{korostelev1993,polonik1995,rigollet2009}. We
will consider a related metric, the volume of the symmetric set difference,
which weights points according to their degree in $G_{n,r}$.
\begin{definition}
  \label{def:volume_symmetric_set_difference}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and a set 
  $\mathcal{S} \subseteq \Rd$, their symmetric set difference is 
  $$
  \Cest \vartriangle \mathcal{S}[\Xbf] :=
    \bigl(\Cest \setminus \mathcal{S}[\Xbf]\bigr) \cup
    \bigl(\mathcal{S}[\Xbf] \setminus \Cest\bigr).
  $$
  Furthermore, we denote the volume of the symmetric set difference by 
  $$
  \Delta(\Cest, \mathcal{S}[\Xbf]) := \vol(\Cest \vartriangle \mathcal{S}[\Xbf];
  G_{n,r}). 
  $$
\end{definition}

@@

These results are of fundamental importance; however, the behavior of the
spectra of these continuum operators can in general be hard to grasp. Therefore,
further work relating this spectra to the geometry of the unlerlying
distribution $\Pbb$ is of interest. In this spirit,
\citet{shi2009,schiebinger2015,garciatrillos19} examine the ability of spectral
algorithms to recover the latent labels in certain geometrically
well-conditioned nonparametric mixture models. Their results focus on global
rather than local methods, and thus impose global rather than local conditions
on the nature of the density. Moreover, they do not in general guarantee
recovery of density clusters, which is the focus in our work. Perhaps most
importantly, these works rely on general cluster saliency conditions, which
implicitly depend on many distinct geometric aspects of the cluster $\Cset$
under consideration. We make this dependence more explicit, and in doing so
expose the role each geometric condition plays in the clustering problem.

Our analysis naturally builds on a few of the aforementioned theoretical
analyses of PPR. For an arbitrary graph $G$ and subset $S \subseteq G$,
\citet{andersen2006} relate the quality of the PPR cluster \smash{$\Cest$} to
the normalized cut functional $\Phi(S;G)$. While this analysis is tight in a
worst-case sense, it fails to account for possible improvements when the cluster
$S$ is additionally assumed to be internally well-connected, which is an
intuitively more favorable case for clustering. Building on this intuition,  
\citet{zhu2013} assume that the subgraph $G[S]$ is internally
well-connected---as measured by a functional such as mixing time of a random
walk over $G[S]$---and prove upper bounds on \smash{$\vol(\Cest \vartriangle
  S;G)$}. Both of these analyses also hold with respect to an approximate form 
of PPR  (APPR), which can be efficiently computed.

@@

To motivate these conditions, we now give a brief high-level sketch
of our analysis (which we will return to more formally in Section~\ref{sec:  
  analysis}).  \citet{zhu2013} show that for an arbitrary graph $G = (V,E)$ and
subset of vertices $S \subseteq V$, the PPR algorithm (properly initialized
within $S$) will output an estimate \smash{$\Cest$} of $S$ satisfying, for a
constant $M>0$, 
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \vartriangle S; G) \leq M \cdot \Phi(S,G) \cdot \tau_{\infty}(G[S]),    
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is  the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn: graph_symmetric_set_difference_1} is one of our principle metrics
of interest, the volume of the symmetric set difference, and our main goal will 
therefore be to upper bound the graph functionals $\Phi$ and $\tau_{\infty}$.
Towards this goal, as we will show in Section
\ref{sec: analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp: 
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. Specifically, assumption 
\ref{asmp: low_noise_density} yields an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density} and \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.

@@

\begin{theorem}
  \label{thm: volume_ssd}
  Fix $\lambda,\sigma > 0$, let \smash{$\Cset \in \Cbb_f(\lambda)$} be a   
  $\kappa$-well-conditioned density cluster, and assume Algorithm \ref{alg: ppr}
  is well-initialized with respect to $\Cset$. There exists a set $\Csig[\Xbf]^g 
  \subseteq \Csig[\Xbf]$ of large volume, \smash{$\vol(\Csig[\Xbf]^g; G_{n,r})
    \geq \vol(\Csig[\Xbf];G_{n,r})/2$}, such that the following holds: if 
  Algorithm~\ref{alg: ppr} is run with any seed node $v \in \Csig[\Xbf]^g$, then
  for 
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  the PPR estimated cluster \smash{$\Cest$} satisfies
  \begin{equation}
    \label{eqn: volume_ssd}
    \Delta(\Cest, \Csig[\Xbf]) \leq c \cdot \kappa(\Cset) \cdot
    \vol_{n,r}(\Csig[\Xbf]),  
  \end{equation}
  with probability at least $1 - b_2/n$. 
\end{theorem}

@@

The bound on symmetric set difference \eqref{eqn: volume_ssd} does not imply 
consistent density cluster estimation in the sense of \eqref{eqn:
  consistent_density_cluster_recovery}. This notion of consistency requires a
uniform bound over the PPR vector $\pbf_v$: as an example, suppose that we were
able to show that for all \smash{$\Cset' \in \Cbb_f(\lambda), \Cset' \neq
  \Cset$}, and each $u \in \Cset, w \in \Cset'$,  
\begin{equation}
\label{eqn: ppr_gap}
\frac{\pbf_v(w)}{D_{ww}} \leq \frac{1}{100 {n \choose 2} \vol_{\Pbb,r}(\Csig)} <
\frac{1}{10 {n \choose 2} \vol_{\Pbb,r}(\Csig)} \leq \frac{\pbf_v(u)}{D_{uu}}. 
\end{equation}
Then, any $(L,U)$ satisfying~\eqref{eqn: initialization} and any sweep cut
$S_{\beta}$ for $\beta \in (L,U)$ will fulfill both conditions laid out in
\eqref{eqn: consistent_density_cluster_recovery}. In Theorem 
\ref{thm: consistent_recovery_of_density_clusters}, we show that a sufficiently 
small upper bound on $\kappa(\Cset)$ ensures that with high probability the
uniform bound~\eqref{eqn: ppr_gap} is satisfied, and hence implies
\smash{$\Cest$} will be a consistent  estimator. We will need one additional 
regularity condition, to preclude arbitrarily low degree vertices for points $x
\in \Cset'[\Xbf]$.

@@

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Under the assumptions of Theorem \ref{thm: volume_ssd}, additionally assume 
  \ref{asmp: C'_bounded_density}, and 
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{(\lambda_{\sigma} r^d
      \nu_d)^2}{\vol_{\Pbb,r}(\Csig)}.
  \end{equation}
  There exists a set $\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$ of large volume, 
  \smash{$\vol(\Csig[\Xbf]^g; G_{n,r}) \geq
    \frac{1}{2}\vol(\Csig[\Xbf];G_{n,r})$}, such that if Algorithm~\ref{alg:
    ppr} is run with any seed node $v \in \Csig[\Xbf]^g$, then for
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  the PPR estimated cluster \smash{$\Cest$} satisfies \eqref{eqn:
    consistent_density_cluster_recovery} with probability at least $1-b_2/n$.
\end{theorem}

@@

\begin{remark}
  Letting the radius of the neighborhood graph shrink, $r \to 0$ as $n \to  
  \infty$, would be computationally attractive (it would ensure that the graph 
  $G_{n,r}$ is sparse), but the presence of a factor of \smash{$\log^2(1/r)/r$}
  in $\kappa(\Cset)$ prevents us from making claims about the behavior of PPR in
  this regime. Although the restriction to a kernel function fixed in $n$ is
  common in spectral clustering theory \citep{schiebinger2015,vonluxburg2008},
  it is an interesting question whether PPR exhibits some degeneracy over 
  neighborhood graphs such that $r \to 0$, or if this is merely looseness in our  
  upper bounds.
\end{remark}

@@

\begin{corollary}
  \label{cor: appr}
  Under the assumptions of Theorem \ref{thm: volume_ssd}, consider instead of
  Algorithm \ref{alg: ppr}, using the approximate PPR vector from
  \citet{andersen2006} satisfying \eqref{eqn: appr_error}, and forming the 
  corresponding cluster estimate \smash{$\Cest$} in the same manner.  Then 
  provided we take 
  \begin{equation}
    \label{eqn: appr_parameter}
    \varepsilon = \frac{1}{25} \vol_{\Pbb,r}(\Csig),
  \end{equation}
  the upper bound on symmetric set difference in \eqref{eqn: volume_ssd} still
  holds. Further, under the assumptions of Theorem \ref{thm:
    consistent_recovery_of_density_clusters}, the estimator \smash{$\Cest$} is
  still consistent, in the sense of Definition \ref{def:
    consistent_density_cluster_estimation}.  
\end{corollary}

@@

\begin{remark}
  Observe that the diameter $\rho$ is absent from Theorem \ref{thm:
    conductance_upper_bound}, in contrast to ultimate bound in Theorem    
  \ref{thm: volume_ssd} (where the diameter enters through the condition number 
  $\kappa(\Cset)$, which worsens (increases) as $\rho$ increases. This reflects
  (what may be regarded as) established wisdom regarding spectral partitioning
  algorithms more generally \citep{guattery1995, hein2010}, albeit newly applied
  to the density clustering setting: if the diameter $\rho$ is large, then PPR
  may fail to recover $\Csig[\Xbf]$ even when $\Cset$ is sufficiently
  well-conditioned to ensure that $\Csig[\Xbf]$ has a small normalized cut in 
  $G_{n,r}$. This will be supported by simulations in Section \ref{sec:
    experiments}.   
\end{remark}

@@

\begin{theorem}
  \label{thm: mixing_time_upper_bound}
  Fix $\lambda, \sigma > 0$, and assume \smash{$\Cset \in \Cbb_f(\lambda)$} 
  satisfies Assumptions \ref{asmp: bounded_density} and \ref{asmp: embedding}.
  Also assume that \smash{$0 < r \leq \sigma/(2\sqrt{d})$}. Then for
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  the mixing time satisfies
  \begin{equation} 
    \label{eqn: mixing_time_upper_bound}
    \tau_{\infty}(G_{n,r}[\Csig[\Xbf]]) \leq c_2 \frac{\Lambda_{\sigma}^4 d^3 
      \rho^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{
        \Lambda_{\sigma}}{\lambda_{\sigma}^2 r}\right) + c_3, 
  \end{equation}
  with probability at least $1 - b_2/n$.
\end{theorem}
The proof of Theorem \ref{thm: mixing_time_upper_bound} relies heavily on
analogous mixing time bounds developed for a continuous-space ``ball walk'' over
convex sets. To the best of our knowledge, our result is the first bound on the
mixing time of random walks over neighborhood graphs that is independent of $n$,
the number of vertices.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.495\textwidth]{lower_bound_plots/lower_bound_1}
  \includegraphics[width=0.495\textwidth]{lower_bound_plots/lower_bound_2}
  \caption{\it\small The density $f$ in \eqref{eqn:lb_density}, for
    $\rho=1$, and two different choices of $\epsilon$ and $\sigma$. Left:
    $\epsilon = 0.3$ and $\sigma = 0.1$; right: $\epsilon = 0.2$ and 
    $\sigma = 0.2$.} 
  \label{fig: hard_case}
\end{figure}

@@

\begin{theorem}
  \label{thm:ppr_lb}
  Assume Algorithm~\ref{alg: ppr} is initialized using inputs $r < 
  \min\set{\frac{1}{40}\rho, \frac{1}{4}\sigma}$, $\alpha = 65  
  \Phi_{\Pbb}(\mathcal{L})$, and $(L,U) = (0,1)$.  Then, for any 
  $$
  n \geq \max\set{\frac{64}{\epsilon^2 \rho \sigma \pi r^2},
    \frac{8}{\epsilon}}, 
  $$
  there exists a set $\Cset[X]^g$ of large volume, \smash{$\vol(\Cset[X]^g \cap 
    \Cset^{(1)}[\Xbf]) \geq \frac{1}{10}\vol_{n,r}(\Cset^{(1)}[\Xbf])$}, such
  that for any seed node $v \in \Cset[X]^g$, the PPR estimated cluster
  \smash{$\Cest$} satisfies    
  \begin{equation}
    \label{eqn:ppr_lb}
    \frac{\sigma \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle
    \Cset^{(1)}[\Xbf]) \geq \frac{1}{4} -  c
    \frac{\sqrt{\sigma/\rho}}{\epsilon^2} \sqrt{ \log\left(\frac{\rho \sigma}
        {\epsilon^2 r^2}\right) \frac{\sigma}{r}},   
  \end{equation}
  with probability at least $1 - b_1 n \exp\set{-b_2 n}$.  Consequently, if
  $$
  \epsilon^2 > \frac{c}{8} \sqrt{\frac{\sigma}{\rho}} \cdot \sqrt{ 
    \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}, 
  $$
  then with high probability \smash{$\frac{\sigma
    \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle \Cset^{(1)}[\Xbf]) 
  \geq 1/8$}.    
\end{theorem}

Note that $\vol_{n,r}(\Cset^{(1)})$ for large enough $n$ will be of the order
$n^2 r^2/(\sigma\rho)$, and therefore the quantity \smash{$\frac{\sigma
    \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle \Cset^{(1)}[\Xbf])$} in 
\eqref{eqn:ppr_lb} is comparable to \smash{$\vol_{n,r}(\Cest
  \vartriangle \Cset^{(1)}[\Xbf]) / \vol_{n,r}(\Cset^{(1)}[\Xbf])$}, which
corresponds to the quantity we upper bound in Theorem~\ref{thm: volume_ssd}.

>>>>

\begin{abstract}
We analyze the Personalized PageRank (PPR) algorithm, a local spectral method
for clustering, which extracts clusters using locally-biased random walks around
a given seed node.  In contrast to previous work, we adopt a classical
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters, that satisfy a set of natural geometric
conditions. We also show a converse result, that PPR can fail to recover
geometrically poorly-conditioned density clusters, even asymptotically. Finally,
we provide empirical support for our theory.
\end{abstract}
\end{center}

@@

First, we define the PPR vector $\pbf_v = \pbf(v,\alpha;G_{n,r})$, based on 
a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to
be the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf_v = \alpha \ebf{v} + (1 - \alpha) \pbf_v \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over
$G_{n,r}$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in
position $v$ and 0 elsewhere).

@@

Let \smash{$\Cbb_f(\lambda)$} denote the connected components of the density
upper level set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster
\smash{$\Cset \in \Cbb_f(\lambda)$}, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the
\emph{empirical density cluster}. The size of the symmetric set difference
between estimated and empirical cluster is a commonly used metric to quantify
cluster estimation error \citep{korostelev1993,polonik1995,rigollet2009}. We
will consider a related metric, the volume of the symmetric set difference,
which weights points according to their degree in $G_{n,r}$. To keep things simple, for a given set $S \subseteq \Xbf$ we write $\vol_{n,r}(S) := \vol(S;G_{n,r})$. 
\begin{definition}
  \label{def:volume_symmetric_set_difference}
  For an estimator \smash{$\Cest \subseteq \Xbf$} and a set 
  $\mathcal{S} \subseteq \Rd$, their symmetric set difference is 
  $$
  \Cest \vartriangle \mathcal{S}[\Xbf] :=
    \bigl(\Cest \setminus \mathcal{S}[\Xbf]\bigr) \cup
    \bigl(\mathcal{S}[\Xbf] \setminus \Cest\bigr).
  $$
  Furthermore, we denote the volume of the symmetric set difference by 
  $$
  \Delta(\Cest, \mathcal{S}[\Xbf]) := \vol_{n,r}(\Cest \vartriangle \mathcal{S}[\Xbf]). 
  $$
\end{definition}

@@

These results are of fundamental importance; however, the behavior of the
spectra of these continuum operators can in general be hard to grasp. Therefore,
further work relating this spectra to the geometry of the underlying
distribution $\Pbb$ is of interest. In this spirit,
\citet{shi2009,schiebinger2015,garciatrillos19} examine the ability of spectral
algorithms to recover the latent labels in certain geometrically
well-conditioned nonparametric mixture models. Their results focus on global
rather than local methods, and thus impose global rather than local conditions
on the nature of the density. Moreover, they do not in general guarantee
recovery of density clusters, which is the focus in our work. Perhaps most
importantly, these works rely on general cluster saliency conditions, which
implicitly depend on many distinct geometric aspects of the cluster $\Cset$
under consideration. We make this dependence more explicit, and in doing so
expose the role each geometric condition plays in the clustering problem.

Our analysis naturally builds on a few of the aforementioned theoretical
analyses of PPR. For an arbitrary graph $G$ and subset $S \subseteq G$,
\citet{andersen2006} relate the quality of the PPR cluster \smash{$\Cest$} to
the normalized cut functional $\Phi(S;G)$. While this analysis is tight in a
worst-case sense, it fails to account for possible improvements when the cluster
$S$ is additionally assumed to be internally well-connected, which is an
intuitively more favorable case for clustering. Building on this intuition,  
\citet{zhu2013} assume that the subgraph $G[S]$ is internally
well-connected---as measured by a functional such as mixing time of a random
walk over $G[S]$---and prove upper bounds on \smash{$\vol(\Cest \vartriangle
  S;G)$}. Both of these analyses also hold with respect to an approximate form 
of PPR  (aPPR), which can be efficiently computed.

@@

To motivate these conditions, we now give a brief high-level sketch
of our analysis (which we will return to more formally in Section~\ref{sec:  
  analysis}).  \citet{zhu2013} show that for an arbitrary graph $G = (V,E)$ and
subset of vertices $S \subseteq V$, the PPR algorithm (properly initialized
within $S$) will output an estimate \smash{$\Cest$} of $S$ satisfying, for a
constant $M>0$, 
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \vartriangle S; G) \leq M \cdot \Phi(S,G) \cdot \tau_{\infty}(G[S]),    
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is  the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn: graph_symmetric_set_difference_1} is one of our principle metrics
of interest, the volume of the symmetric set difference, and our main goal will 
therefore be to upper bound the graph functionals $\Phi$ and $\tau_{\infty}$.
Towards this goal, as we will show in Section
\ref{sec: analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp: 
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. Specifically, assumption 
\ref{asmp: low_noise_density} yields an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol_{n,r}(\Csig[\Xbf])$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density} and \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.

@@

\begin{theorem}
  \label{thm: volume_ssd}
  Fix $\lambda,\sigma > 0$, let \smash{$\Cset \in \Cbb_f(\lambda)$} be a   
  $\kappa$-well-conditioned density cluster, and assume Algorithm \ref{alg: ppr}
  is well-initialized with respect to $\Cset$. Then for any 
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  there exists a set $\Csig[\Xbf]^g 
  \subseteq \Csig[\Xbf]$ of large volume, \smash{$\vol_{n,r}(\Csig[\Xbf]^g)
    \geq \vol_{n,r}(\Csig[\Xbf])/2$}, such that the following holds: if 
  Algorithm~\ref{alg: ppr} is run with any seed node $v \in \Csig[\Xbf]^g$, then
  the PPR estimated cluster \smash{$\Cest$} satisfies
  \begin{equation}
    \label{eqn: volume_ssd}
    \Delta(\Cest, \Csig[\Xbf]) \leq c \cdot \kappa(\Cset) \cdot
    \vol_{n,r}(\Csig[\Xbf]),  
  \end{equation}
  with probability at least $1 - b_2/n$. 
\end{theorem}

@@

The bound on symmetric set difference \eqref{eqn: volume_ssd} does not imply 
consistent density cluster estimation in the sense of \eqref{eqn:
  consistent_density_cluster_recovery}. This notion of consistency requires a
uniform bound over the PPR vector $\pbf_v$: as an example, suppose that we were
able to show that for all \smash{$\Cset' \in \Cbb_f(\lambda), \Cset' \neq
  \Cset$}, and each $u \in \Cset, w \in \Cset'$,  
\begin{equation}
\label{eqn: ppr_gap}
\frac{\pbf_v(w)}{D_{ww}} \leq \frac{1}{100 {n \choose 2} \vol_{\Pbb,r}(\Csig)} <
\frac{1}{10 {n \choose 2} \vol_{\Pbb,r}(\Csig)} \leq \frac{\pbf_v(u)}{D_{uu}}. 
\end{equation}
Then, any $(L,U)$ satisfying~\eqref{eqn: initialization} and any sweep cut
$S_{\beta}$ for $\beta \in (L,U)$ would fulfill both conditions laid out in
\eqref{eqn: consistent_density_cluster_recovery}. In Theorem 
\ref{thm: consistent_recovery_of_density_clusters}, we show that a sufficiently 
small upper bound on $\kappa(\Cset)$ ensures that with high probability the
uniform bound~\eqref{eqn: ppr_gap} is satisfied, and hence implies
\smash{$\Cest$} will be a consistent  estimator. We will need one additional 
regularity condition, to preclude arbitrarily low degree vertices for points $x
\in \Cset'[\Xbf]$.

@@

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Under the assumptions of Theorem \ref{thm: volume_ssd}, additionally assume 
  \ref{asmp: C'_bounded_density}, and 
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c \frac{(\lambda_{\sigma} r^d
      \nu_d)^2}{\vol_{\Pbb,r}(\Csig)}.
  \end{equation}
  Then for any
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  there exists a set $\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$ of large volume, 
  \smash{$\vol_{n,r}(\Csig[\Xbf]^g) \geq
    \vol_{n,r}(\Csig[\Xbf])/2$}, such that if Algorithm~\ref{alg:
    ppr} is run with any seed node $v \in \Csig[\Xbf]^g$, then the PPR estimated cluster \smash{$\Cest$} satisfies \eqref{eqn: consistent_density_cluster_recovery} with probability at least $1-b_2/n$.
\end{theorem}

@@

\begin{remark}
  Letting the radius of the neighborhood graph shrink, $r \to 0$ as $n \to  
  \infty$, would be computationally attractive (it would ensure that the graph 
  $G_{n,r}$ is sparse), but the presence of a factor of \smash{$\log^2(1/r)/r$}
  in $\kappa(\Cset)$ prevents us from making claims about the behavior of PPR in
  this regime. Although the restriction to a kernel function fixed in $n$ is
  common in spectral clustering theory \citep{schiebinger2015,vonluxburg2008},
  it is an interesting question whether PPR exhibits some degeneracy over 
  neighborhood graphs as $r \to 0$, or if this is merely looseness in our  
  upper bounds.
\end{remark}

@@

\begin{corollary}
  \label{cor: appr}
  Consider instead of
  Algorithm \ref{alg: ppr} using the approximate PPR vector from
  \citet{andersen2006} satisfying \eqref{eqn: appr_error}, and forming the 
  corresponding cluster estimate \smash{$\Cest$} in the same manner.  Then 
  provided we take 
  \begin{equation}
    \label{eqn: appr_parameter}
    \varepsilon = \frac{1}{25} \vol_{\Pbb,r}(\Csig),
  \end{equation}
  under the assumptions of Theorem~\ref{thm: volume_ssd} the upper bound on symmetric set difference in \eqref{eqn: volume_ssd} still
  holds, and under the assumptions of Theorem \ref{thm: consistent_recovery_of_density_clusters} the set inclusion and disjointedness statements in \eqref{eqn: consistent_density_cluster_recovery} still hold, each with probability at least $1 - b_2/n$ (under possibly different choices of the constants $c_i$ and $b_i$).
\end{corollary}

@@

\begin{remark}
  Observe that the diameter $\rho$ is absent from Theorem \ref{thm:
    conductance_upper_bound}, in contrast to the ultimate bound in Theorem    
  \ref{thm: volume_ssd} where the diameter enters through the condition number 
  $\kappa(\Cset)$, which worsens (increases) as $\rho$ increases. This reflects
  (what may be regarded as) established wisdom regarding spectral partitioning
  algorithms more generally \citep{guattery1995, hein2010}, albeit newly applied
  to the density clustering setting: if the diameter $\rho$ is large, then PPR
  may fail to recover $\Csig[\Xbf]$ even when $\Cset$ is sufficiently
  well-conditioned to ensure that $\Csig[\Xbf]$ has a small normalized cut in 
  $G_{n,r}$. This will be supported by simulations in Section \ref{sec:
    experiments}.   
\end{remark}

@@

\begin{theorem}
  \label{thm: mixing_time_upper_bound}
  Fix $\lambda, \sigma > 0$, and assume \smash{$\Cset \in \Cbb_f(\lambda)$} 
  satisfies Assumptions \ref{asmp: bounded_density} and \ref{asmp: embedding}.
  Also assume that \smash{$0 < r \leq \sigma/(2\sqrt{d})$}. Then for any
  $$
  n \geq b_1 (\log n)^{\max\{\frac{3}{d},1\}},
  $$
  the mixing time satisfies
  \begin{equation} 
    \label{eqn: mixing_time_upper_bound}
    \tau_{\infty}(G_{n,r}[\Csig[\Xbf]]) \leq c_2 \frac{\Lambda_{\sigma}^4 d^3 
      \rho^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{
        \Lambda_{\sigma}}{\lambda_{\sigma}^2 r}\right) + c_3, 
  \end{equation}
  with probability at least $1 - b_2/n$.
\end{theorem}
The proof of Theorem \ref{thm: mixing_time_upper_bound} relies heavily on
analogous mixing time bounds developed for a continuous-space ``ball walk'' over
convex sets. To the best of our knowledge, our result is the first bound on the
mixing time of random walks over neighborhood graphs that is independent of $n$,
the number of vertices.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.495\textwidth]{lower_bound_1}
  \includegraphics[width=0.495\textwidth]{lower_bound_2}
  \caption{\it\small The density $f$ in \eqref{eqn:lb_density}, for
    $\rho=1$, and two different choices of $\epsilon$ and $\sigma$. Left:
    $\epsilon = 0.3$ and $\sigma = 0.1$; right: $\epsilon = 0.2$ and 
    $\sigma = 0.2$.} 
  \label{fig: hard_case}
\end{figure}

@@

\begin{theorem}
  \label{thm:ppr_lb}
  Assume Algorithm~\ref{alg: ppr} is initialized using inputs $r < 
  \min\set{\frac{1}{40}\rho, \frac{1}{4}\sigma}$, $\alpha = 65  
  \Phi_{\Pbb}(\mathcal{L})$, and $(L,U) = (0,1)$.  Then, for any 
  $$
  n \geq \max\set{\frac{64}{\epsilon^2 \rho \sigma \pi r^2},
    \frac{8}{\epsilon}}, 
  $$
  there exists a set $\Cset[X]^g$ of large volume, \smash{$\vol_{n,r}(\Cset[X]^g \cap 
    \Cset^{(1)}[\Xbf]) \geq \vol_{n,r}(\Cset^{(1)}[\Xbf];G_{n,r})/10$}, such
  that for any seed node $v \in \Cset[X]^g$, the PPR estimated cluster
  \smash{$\Cest$} satisfies    
  \begin{equation}
    \label{eqn:ppr_lb}
    \frac{\sigma \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle
    \Cset^{(1)}[\Xbf]) \geq \frac{1}{4} -  c
    \frac{\sqrt{\sigma/\rho}}{\epsilon^2} \sqrt{ \log\left(\frac{\rho \sigma}
        {\epsilon^2 r^2}\right) \frac{\sigma}{r}},   
  \end{equation}
  with probability at least $1 - b_1 n \exp\set{-b_2 n}$.  Consequently, if
  $$
  \epsilon^2 > \frac{c}{8} \sqrt{\frac{\sigma}{\rho}} \cdot \sqrt{ 
    \log\left(\frac{\rho \sigma}{\epsilon^2 r^2}\right)\frac{\sigma}{r}}, 
  $$
  then with high probability \smash{$\frac{\sigma
    \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle \Cset^{(1)}[\Xbf]) 
  \geq 1/8$}.    
\end{theorem}

Note that $\vol_{n,r}(\Cset^{(1)}[\Xbf])$ for large enough $n$ will be of the order
$n^2 r^2/(\sigma\rho)$, and therefore the quantity \smash{$\frac{\sigma
    \rho}{r^2 n^2} \cdot \vol_{n,r}(\Cest \vartriangle \Cset^{(1)}[\Xbf])$} in 
\eqref{eqn:ppr_lb} is comparable to \smash{$\vol_{n,r}(\Cest
  \vartriangle \Cset^{(1)}[\Xbf]) / \vol_{n,r}(\Cset^{(1)}[\Xbf])$}, which
corresponds to the quantity we upper bound in Theorem~\ref{thm: volume_ssd}.

Theorem~\ref{thm:ppr_lb} is stated with respect to a particular hard case, where
the density clusters are rectangular subsets of $\Reals^2$.  We chose this
setting to make the theorem simple to state, and our results are generalizable
to $\Reals^d$ and to non-rectangular clusters.  Moreover, although we state
our lower bound with respect to PPR run on a neighborhood graph, the conclusion is
likely to hold for a much broader class of spectral clustering algorithms. In
the proof of Theorem~\ref{thm:ppr_lb}, we rely heavily on the fact that when
$\epsilon^2$ is sufficiently greater than $\sigma/\rho$, the normalized cut of
$\Cset^{(1)}$ will be much larger than that of $\mathcal{L}$. In this case, not
merely PPR but any algorithm that approximates the minimum normalized cut is
unlikely to recover $\Cset^{(1)}$. In particular, local spectral clustering
algorithms based on truncated random walks \citet{spielman2013}, global spectral
clustering algorithms \citet{shi00}, and $p$-Laplacian based spectral embeddings
\citet{hein2010} all have provable upper bounds on the normalized cut of cluster
they output, and thus we expect that they would all fail to estimate
$\Cset^{(1)}$.
